Automatic Differentiation Accelerates Inference for Partially Observed Markov Processes

### 2024-08-20 paper/ms.tex

However, \cite{scibior21} used their estimator within an algorithm which does not have the plug-and-play property and which has high Monte Carlo variability, while the stop-gradient procedure is introduced without mathematical motivation.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.

@@

The result in Theorem \ref{thm:mop-functional-forms} provides mathematical intuition for why \cite{scibior21}'s use of the stop-gradient procedure in the particle filter recovers the estimate of \cite{poyiadjis11}. When $\theta=\phi$, the correction $g_{n,k_j}^\theta/g_{n,k_j}^\phi$ is equivalent to $g_{n,k_j}^\theta / \text{stop\_gradient}(g_{n,k_j}^\theta)$. As such, the use of the stop-gradient in \cite{scibior21} can be seen as a special case of the MOP-$\alpha$ weight correction for $\alpha=1, \theta=\phi$.

@@

\arxiv{}{\vspace*{-4mm}}
\section{Practical Maximum-Likelihood Estimation}

%One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.

>>>>

However, the algorithm of \cite{scibior21} that can be modified to have the plug-and-play property recovers an estimator with high Monte Carlo variability \cite{poyiadjis11}. Additionally, while the stop-gradient procedure is undeniably effective, the reason why the correction recovers the \cite{poyiadjis11} estimator remains somewhat unclear.
We use \cite{scibior21} and \cite{poyiadjis11} as our starting point in developing a new approach remedying these issues.

@@

The result in Theorem \ref{thm:mop-functional-forms} provides mathematical intuition for why \cite{scibior21}'s use of the stop-gradient procedure in the particle filter recovers the estimate of \cite{poyiadjis11}. When $\theta=\phi$, the correction $g_{n,k_j}^\theta/g_{n,k_j}^\phi$ is equivalent to $g_{n,k_j}^\theta / \text{stop\_gradient}(g_{n,k_j}^\theta)$, and is implemented in this way in practice. As such, the use of the stop-gradient in \cite{scibior21} can be seen as the MOP-$\alpha$ weight correction for $\alpha=1, \theta=\phi$.

@@

\arxiv{}{\vspace*{-3mm}}
\section{Practical Maximum-Likelihood Estimation}
\arxiv{}{\vspace*{-1mm}}
%One can think of MOP-$\alpha$ as taking a likelihood estimate at $\phi$, reweighting it to instead target the likelihood at $\theta$, and then taking the derivative with respect to $\theta$ of the above estimate.
